---
title: ESS330 Final Project
authors:
  - name: Mel Balcarcel Arias
    affiliation: Colorado State University
    roles: writing
    corresponding: true
  - name: Sean Walsh
    affiliation: Colorado State University
    roles: writing
    corresponding: false
bibliography: references.bib
csl: apa.csl
---

# Introduction, Background, and Motivation
hi



# Data Exploration 
```{r}
EPA_air_quality <- read.csv("conreport2021.csv")
asthma_prevalence <- read.csv("CDC_asthma_2021.csv")
asthma_mort <- read.csv("CDC_asthma_mort2021.csv")
covid_data3 <- read.csv("covid_confirmed_usafacts.csv")
```

# This dataset comes from the EPA and it shows air quality in 2021. Out of the many variables, we may only focus on a few, with a big focus on PM2.5 which can be detrimental to human health. There are a good amount of blank entries that we'll have to account for. 
```{r}
visdat::vis_dat(EPA_air_quality)
```
## This is asthma prevalence on a state by state level in 2021 pulled from the CDC, the data set is pretty complete with a few NA's. 
```{r}
visdat::vis_dat(asthma_prevalence)
```
## This is the asthma mortality rate on a state by state level in 2021, also pulled from the CDC. 
```{r}
visdat::vis_dat(asthma_mort)
```
## For the covid side we decided to go with case count and while this data set covers a large period of time, we will only focus on 2021. There's issues with the date formatting and it creates a lot of issues especially with visualization so the usual vis_dat is not useful here.


# Methods Identification
* From our data, we feel like we have a lot of air quality parameters to work with which can be very useful as we can explore multiple factors at once. We may already have all the data that we need, but there's a chance that could look at other respiratory illnesses in addition to asthma. 
* A big challenge is the covid data because it's just so robust. However, with a lot of cleaning, it could work out. In addition, understanding some of the air quality indices may be challenging but if we focus on 1 or 2, it should be no big issue. 







